Estimating Sleep Quality Metrics Using Free-Living Accelerometer Data From Thigh-Worn Devices in Comparison to an EEG-Based Sleep Tracking Device

### 2023-07-03 manuscript/manuscript.qmd
all stats for ML models calculated. Now stats for biLSTM.

@@

5.  Wake After Sleep Onset (WASO) - This refers to the time spent awake
    after initially falling asleep and before the final awakening. In
    our analysis, a period is counted as 'awake' only if it consists of
    3 or more contiguous 30-second epochs.

@@

The children children included in the analyses were 9.4 years old (SD =
2.1). The number of sleep epochs recorded in the study was 2,035,261,
which accounted for approximately 86% of the total sleep duration. This
metric indicates the granularity of sleep measurement, capturing the
discrete moments of sleep throughout the sleep period. Sleep epochs are
essential for analyzing sleep patterns and stages, as well as for
assessing sleep quality and efficiency.

As reported in @tbl-zm_overview the sleep quality metrics derived from
ZM predictions were modified by the implementation of 5-minute and
10-minute median filters. The Sleep Period Time (SPT) were consistent
across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as this
corresponds to the length of the ZM recording. Total Sleep Time (TST)
and Sleep Efficiency (SE) increased in the filtered data, implying the
filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. Latency
to Persistent Sleep (LPS) also elevated, suggesting the filter smooths
out brief awakenings at sleep onset, leading to a prolonged time to
persistent sleep. The most significant change was seen in Wake After
Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
10-minute filtered data, respectively. The results underscore the role
of median filters on sleep metrics.

```{r}
#| echo: false
#| message: false
#| label: tbl-zm_overview
#| tbl-cap: Overview of characteristics of the ZM sleep quality summaries. Values are represented as mean (SD).

@@

The performance metrics of the biLSTM were...

>>>>

something on sequence length and step size

@@

5.  Wake After Sleep Onset (WASO) - This refers to the time spent awake
    after initially falling asleep and before the final awakening. In
    our analysis, a period is counted as 'awake' only if it consists of
    3 or more contiguous 30-second epochs which is also how the ZM
    summarizes WASO.

@@

The analysis included children with an average age of 9.4 years (SD =
2.1). The raw ZM predictions covered 2,035,261 sleep epochs, or about
86% of the total recording duration.

As reported in @tbl-zm_overview the sleep quality metrics derived from
ZM predictions were modified by the implementation of 5-minute and
10-minute median filters. The Sleep Period Time (SPT) were consistent
across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as this
corresponds to the length of the ZM recording. Total Sleep Time (TST)
and Sleep Efficiency (SE) increased in the filtered data, implying the
filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0% to 86.4 ± 12.7% and 87.5 ± 12.9% respectively. Latency
to Persistent Sleep (LPS) also elevated, suggesting the filter smooths
out brief awakenings at sleep onset, leading to a prolonged time to
persistent sleep. The most significant change was seen in Wake After
Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
10-minute filtered data, respectively. The number of awakenings was also
considerably reduced with the application of filters. In the raw data,
the average number of awakenings was 34.46 ± 11.33 per night, which
reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
filtered data sets respectively. These results underscore the role of
median filters on sleep metrics.

```{r}
#| echo: false
#| message: false
#| label: tbl-zm_overview
#| tbl-cap: Overview of characteristics of the ZM sleep quality summaries per night. Values are represented as mean (SD).

@@

@tbl-biLSTM_performance presents the performance of the three-class
biLSTM multiclassifier on raw and median-filtered (5-minute and
10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores ranging
from 71.36% to 76.04%, indicating overall good performance. Applying
5-minute median filtering improves the metrics further, resulting in F1
Scores ranging from 75.99% to 78.53%, demonstrating enhanced precision
and sensitivity. However, Specificity values are not available for this
configuration. With 10-minute median filtering, F1 Scores range from
73.45% to 73.93%, maintaining good performance but with limited
information on Specificity. Further analysis is required to assess
performance across all classes.

```{r}
#| echo: false
#| message: false
#| tbl-colwidths: auto
#| label: tbl-biLSTM_performance
#| tbl-cap: Performance of the three-class biLSTM multicclassifier.
source("code/table_biLSTM_performance.R")

tbl_biLSTM_performance
```

## Evaluation of Sleep Quality Summaries

BA plots

pearson regression

scatterplots with identity line and best linear fit

### 2023-07-03 manuscript/manuscript.tex
all stats for ML models calculated. Now stats for biLSTM.

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality metrics
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sleep Period Time (SPT) - This refers to the total duration of the
  sleep period, which is defined as the time from the start to the end
  of the ZM recording.
\item
  Total Sleep Time (TST) - This is the time spent asleep within the SPT.
\item
  Sleep Efficiency (SE) - This is the ratio between TST and SPT,
  representing the proportion of the sleep period that was actually
  spent asleep.
\item
  Latency Until Persistent Sleep (LPS) - This metric represents the time
  it takes to transition from wakefulness to sustained sleep. It is
  calculated as the time from the beginning of the ZM recording until a
  period when 10 out of 12 minutes are scored as sleep.
\item
  Wake After Sleep Onset (WASO) - This refers to the time spent awake
  after initially falling asleep and before the final awakening. In our
  analysis, a period is counted as `awake' only if it consists of 3 or
  more contiguous 30-second epochs.
\end{enumerate}

@@

The children children included in the analyses were 9.4 years old (SD =
2.1). The number of sleep epochs recorded in the study was 2,035,261,
which accounted for approximately 86\% of the total sleep duration. This
metric indicates the granularity of sleep measurement, capturing the
discrete moments of sleep throughout the sleep period. Sleep epochs are
essential for analyzing sleep patterns and stages, as well as for
assessing sleep quality and efficiency.

As reported in Table~\ref{tbl-zm_overview} the sleep quality metrics
derived from ZM predictions were modified by the implementation of
5-minute and 10-minute median filters. The Sleep Period Time (SPT) were
consistent across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as
this corresponds to the length of the ZM recording. Total Sleep Time
(TST) and Sleep Efficiency (SE) increased in the filtered data, implying
the filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ± 12.9\% respectively.
Latency to Persistent Sleep (LPS) also elevated, suggesting the filter
smooths out brief awakenings at sleep onset, leading to a prolonged time
to persistent sleep. The most significant change was seen in Wake After
Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
10-minute filtered data, respectively. The results underscore the role
of median filters on sleep metrics.

\hypertarget{tbl-zm_overview}{}
\begin{longtable}{lrrrrrr}
\caption{\label{tbl-zm_overview}Overview of characteristics of the ZM sleep quality summaries. Values
are represented as mean (SD). }\tabularnewline

@@

\toprule
model & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
\midrule
Decision Tree & $94.36$ & $95.27$ & $93.12$ & $95.64$ & $96.86$ \\ 
Logistic Regression & $94.99$ & $95.74$ & $95.03$ & $94.94$ & $96.26$ \\ 
Feed-Forward Neural Net & $95.03$ & $95.77$ & $95.07$ & $94.99$ & $96.29$ \\ 
XGBoost & $95.38$ & $96.06$ & $95.83$ & $94.94$ & $96.23$ \\ 
\bottomrule
\end{longtable}

@@

The performance metrics of the biLSTM were\ldots{}

>>>>

\begin{keyword}
    Sleep \sep Accelerometry \sep EEG \sep Machine learning \sep 
    Sleep quality metrics
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sleep Period Time (SPT) - This refers to the total duration of the
  sleep period, which is defined as the time from the start to the end
  of the ZM recording.
\item
  Total Sleep Time (TST) - This is the time spent asleep within the SPT.
\item
  Sleep Efficiency (SE) - This is the ratio between TST and SPT,
  representing the proportion of the sleep period that was actually
  spent asleep.
\item
  Latency Until Persistent Sleep (LPS) - This metric represents the time
  it takes to transition from wakefulness to sustained sleep. It is
  calculated as the time from the beginning of the ZM recording until a
  period when 10 out of 12 minutes are scored as sleep.
\item
  Wake After Sleep Onset (WASO) - This refers to the time spent awake
  after initially falling asleep and before the final awakening. In our
  analysis, a period is counted as `awake' only if it consists of 3 or
  more contiguous 30-second epochs which is also how the ZM summarizes
  WASO.
\end{enumerate}

@@

The analysis included children with an average age of 9.4 years (SD =
2.1). The raw ZM predictions covered 2,035,261 sleep epochs, or about
86\% of the total recording duration.

As reported in Table~\ref{tbl-zm_overview} the sleep quality metrics
derived from ZM predictions were modified by the implementation of
5-minute and 10-minute median filters. The Sleep Period Time (SPT) were
consistent across raw and filtered datasets (mean: 9.2 ± 2.1 hours) as
this corresponds to the length of the ZM recording. Total Sleep Time
(TST) and Sleep Efficiency (SE) increased in the filtered data, implying
the filters categorize some wakefulness as sleep. Specifically, TST
increased from a raw mean of 7.7 ± 1.9 hours to 8.1 ± 2.0 hours
(5-minute filter) and 8.2 ± 2.1 hours (10-minute filter), while SE rose
from 82.6 ± 12.0\% to 86.4 ± 12.7\% and 87.5 ± 12.9\% respectively.
Latency to Persistent Sleep (LPS) also elevated, suggesting the filter
smooths out brief awakenings at sleep onset, leading to a prolonged time
to persistent sleep. The most significant change was seen in Wake After
Sleep Onset (WASO), which dropped from 39.0 ± 33.6 minutes in raw data
to 30.6 ± 46.8 minutes and 22.3 ± 55.4 minutes in the 5-minute and
10-minute filtered data, respectively. The number of awakenings was also
considerably reduced with the application of filters. In the raw data,
the average number of awakenings was 34.46 ± 11.33 per night, which
reduced to 4.43 ± 3.26 and 1.95 ± 2.01 for the 5-minute and 10-minute
filtered data sets respectively. These results underscore the role of
median filters on sleep metrics.

\hypertarget{tbl-zm_overview}{}
\begin{longtable}{lrrrrrr}
\caption{\label{tbl-zm_overview}Overview of characteristics of the ZM sleep quality summaries per night.
Values are represented as mean (SD). }\tabularnewline

@@

\toprule
 & F1 Score (\%) & Accuracy (\%) & Sensitivity (\%) & Precision (\%) & Specificity (\%) \\ 
\midrule
Decision Tree & $94.36$ & $95.27$ & $93.12$ & $95.64$ & $96.86$ \\ 
Logistic Regression & $94.99$ & $95.74$ & $95.03$ & $94.94$ & $96.26$ \\ 
Feed-Forward Neural Net & $95.03$ & $95.77$ & $95.07$ & $94.99$ & $96.29$ \\ 
XGBoost & $95.38$ & $96.06$ & $95.83$ & $94.94$ & $96.23$ \\ 
\bottomrule
\end{longtable}

@@

Table~\ref{tbl-biLSTM_performance} presents the performance of the
three-class biLSTM multiclassifier on raw and median-filtered (5-minute
and 10-minute) ZM predictions. Raw ZM predictions achieve F1 Scores
ranging from 71.36\% to 76.04\%, indicating overall good performance.
Applying 5-minute median filtering improves the metrics further,
resulting in F1 Scores ranging from 75.99\% to 78.53\%, demonstrating
enhanced precision and sensitivity. However, Specificity values are not
available for this configuration. With 10-minute median filtering, F1
Scores range from 73.45\% to 73.93\%, maintaining good performance but
with limited information on Specificity. Further analysis is required to
assess performance across all classes.

\hypertarget{tbl-biLSTM_performance}{}
\begin{longtable}{cccc}
\caption{\label{tbl-biLSTM_performance}Performance of the three-class biLSTM multicclassifier. }\tabularnewline

\toprule
 & F1 Score & Sensitivity & Precision \\ 
\midrule
Raw ZM Predictions & $71.36$ & $70.42$ & $76.04$ \\ 
5-Min Median & $75.99$ & $74.62$ & $78.53$ \\ 
10-Min Median & $73.45$ & $73.93$ & $73.07$ \\ 
\bottomrule
\end{longtable}

\hypertarget{evaluation-of-sleep-quality-summaries}{%
\subsection{Evaluation of Sleep Quality
Summaries}\label{evaluation-of-sleep-quality-summaries}}

BA plots

pearson regression

scatterplots with identity line and best linear fit
